Modular analysis of gene expression data with R

### 2010-03-15 isanote/answers_to_reviewers.tex

This Application Note presents a software package that implements the
Iterative Signature Algorithm and provides a convenient visualization
of it’s results. The method is useful and the software may be of help
to interested users. 
\end{quote}

\begin{quote}
I have two problems with the Note in its present form.
The first may be due to my misunderstanding of the method. My
understanding was that some threshold Tg is applied on the genes and
another one, Tc, on the conditions. The aim is to reach a bicluster (a
set of conditions C and a set of genes G) such that G contains only
genes with expression above Tg for all the conditions of C, and
conditions are included in C only if the expression of all genes of G
exceeds Tc (one can replace a lower threshold with an upper one). The
heatmap of the Figure implies that my understanding was incorrect –
the conditions satisfy a double threshold (expressions are either
above or below some thresholds) and the genes have expressions above a
Tg in some conditions or below some other Tg’ in other conditions. 
I urge the authors to define what specifies or defines a bicluster –
do it briefly and sacrifice a few lines from elsewhere. 
\end{quote}

Indeed, by default the software uses a double threshold. This can be
changed, as explained in the reference manual of the ISA functions,
and the ISA tutorial titled `The Iterative Signature Algorithm for
Gene Expression Data'.

To  clarify things better, we have added the following to the main
text:
\textsl{The output of ISA is a collection of potentially overlapping
modules. Every module contains genes, that are up- and/or
down-regulated, in samples that belong to the module.
In every module, each gene and each sample is attributed a
score between minus one and one, that reflects the strength of the
association with the module. Moreover, if the scores of two genes of a
module have the same sign, then they are correlated (across the
samples of the module), opposite signs mean
anti-correlation. Similarly, if two sample
scores have the same sign, then these samples are correlated (across
the genes of the module), opposite signs indicate anti-correlation.}

The tutorial on the ISA homepage explains the ISA iteration in detail:
\textsl{The thresholding is an important step of the ISA, without
  thresholding ISA would be equivalent to a (not too effective)
  numerical singular value decomposition (SVD) algorithm. Currently
  thresholding is done by calculating the mean and standard deviation
  of the vector and keeping only elements that are further than a
  given number of standard deviations from the mean. Using the
  ``direction'' parameter, one can keep values that are
  (a) significantly higher (``up''); (b) lower (``down'') than the
  mean; or (c) both (``updown'').}

### 2010-03-15 eisa/inst/doc/EISA_tutorial.Rnw
ISA paper updated for the resubmission.
ISA paper: added Zoltan's changes.

\begin{figure}
\centering
<<condplot-real,fig=TRUE,width=10,height=4,echo=FALSE,results=hide>>=
<<condplot>>
@
\caption{Condition plot for the ALL data and the second module. Every
  sample is represented as a bar, and its ISA score is plotted. The
  scores are also printed at the top or bottom of the bars. The scores
  corresponding to the ISA thresholds are represented as horizontal
  lines. The dashed line is the mean of the condition scores. B-cell
  samples are dark green, T-cell samples are orange.
}
\label{fig:condplot}
\end{figure}

@@

The thresholding is an important step of the ISA, without thresholding
ISA would be equivalent to a (not too effective) numerical singular
value decomposition (SVD) algorithm. Currently thresholding is done by
calculating the mean and standard deviation of the vector and keeping
only elements that are further than a given number of standard
deviations from the mean. Based on the \Rargument{direction} parameter,
this means keeping values that are significantly higher than the
mean (``\emph{up}''), or keeping the ones that are significantly lower
than the mean (``\emph{down}''); or keeping both
(``\emph{updown}'').

>>>>

\begin{figure}
\centering
<<condplot-real,fig=TRUE,width=10,height=4,echo=FALSE,results=hide>>=
<<condplot>>
@
\caption{Condition (or sample score) plot for the ALL data and the
  second module. Every sample is represented as a bar, and its ISA
  score is plotted. The scores are also printed at the top or bottom
  of the bars. The scores corresponding to the ISA thresholds are
  represented as horizontal lines. The dashed line is the mean of the
  condition scores. B-cell samples are dark green, T-cell samples are
  orange. The green and red horizontal lines show the sample
  thresholds. Samples above the red line, and samples below the green
  line are included in the module.
}
\label{fig:condplot}
\end{figure}

@@

The thresholding is an important step of the ISA, without
thresholding ISA would be equivalent to a (not too effective)
numerical singular value decomposition (SVD) algorithm. Currently
thresholding is done by calculating the mean and standard deviation
of the vector and keeping only elements that are further than a
given number of standard deviations from the mean. Using the
``direction'' parameter, one can keep values that are
(a) significantly higher (``up''); (b) lower (``down'') than the
mean; or (c) both (``updown'').

### 2010-03-15 isanote/modular.tex
ISA note: make figure a bit bigger.
ISA paper: added Zoltan's changes.

In this short note, we introduce two BioConductor \citep{BioC} software
packages written in GNU~R: The~\Rpackage{isa2} package includes an optimized
implementation of the ISA and the~\Rpackage{eisa} package provides a
convenient interface 
to run the ISA, visualize its output and put the biclusters into
biological context. Potential users for these packages are all R and
BioConductor users dealing with tabular (e.g. gene expression) data.

@@

The output of ISA is a collection of potentially overlapping
modules. Every module contains genes, that are up- and/or
down-regulated together, exactly in the samples of the module.
In every module, each gene and each sample is attributed a
score between minus one and one, that reflects the strength of the
association with the module. Moreover, if the scores of two genes of a
module have the same sign, then they are correlated (in the samples of
the module), opposite signs mean anti-correlation. If two sample
scores have the same sign, then these samples are correlated (along
the genes of the module), opposite signs signal anti-correlation.

@@

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis with the
  \Rpackage{eisa} package. See text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing coherent
  expression of the genes across the samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} Module tree. Each rectangle is a module; see the
  definition of the edges in the text. Modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the module. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

>>>>

In this short note, we introduce two BioConductor \citep{BioC} software
packages written in GNU~R: The~\Rpackage{isa2} package includes an optimized
implementation of the ISA and the~\Rpackage{eisa} package provides a
convenient interface 
to run the ISA, visualize its output and put the biclusters into
biological context. Potential users of these packages are all R and
BioConductor users dealing with tabular (e.g. gene expression) data.

@@

The output of ISA is a collection of potentially overlapping
modules. Every module contains genes, that are up- and/or
down-regulated, in samples that belong to the module.
In every module, each gene and each sample is attributed a
score between minus one and one, that reflects the strength of the
association with the module. Moreover, if the scores of two genes of a
module have the same sign, then they are correlated (across the
samples of the module), opposite signs mean
anti-correlation. Similarly, if two sample
scores have the same sign, then these samples are correlated (across
the genes of the module), opposite signs indicate anti-correlation.

@@

\begin{figure*}
\centering
\includegraphics[width=0.842\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis with the
  \Rpackage{eisa} package. See text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing coherent
  expression of the genes across the samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} Module tree. Each rectangle is a module; see the
  definition of the edges in the text. Modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the module. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}
